### 2016-01-29 doc/thesis/intro.tex
More writing

\subsection{Phylogenetics and phylodynamics}

\subsubsection{Phylogenetic trees}

@@

The branches of the tree may have associated lengths, representing either
evolutionary distance or calendar time between ancestors and their descendants.
The term ``evolutionary distance'' is used here imprecisely to mean any sort of
quantitative measure of evolution, such as the number of differences between
the DNA sequences of an ancestor its descendant, or the difference in average
body mass or height. This will be clarified below in
subsection~\ref{subsubsec:genetree}. A phylogeny with branch lengths in calendar
time units is often referred to as \defn{time-scaled}. In a time-scaled
phylogeny, the internal nodes can be mapped onto a timeline by using the tips
of the tree, which usually map to the present day, as a reference
point~\autocite{nee1992tempo}. The corresponding points on the timeline are
called \defn{branching times}, and the rate of their accumulation is referred
to as the \defn{branching rate}. Rooted trees whose tips are all the same
distance from the root are called \defn{ultrametric}
trees~\autocite{buneman1974note}. These concepts are illustrated in
Figure~\ref{fig:speciestree}.

@@

\subsubsection{Species trees and transmission trees}
\label{subsubsec:speciestree}

A \defn{species tree} is a particular type of phylogeny in which the taxa are
species, and the branching times correspond to historical speciation events. By
speciation events, we mean times when two formerly interbreeding groups of
organisms ceased to reproduce with each other. In general, we do not have
access to the true tree relating a particular group of extant species, as this
would imply certain knowledge of speciation events in the distant past.
Therefore, all species trees considered by researchers are estimates, based on
the genetic or phenotypic similarity of the extant taxa, the fossil record, or
both. We discuss some of the challenges associated with this estimation in
subsection \ref{subsubsec:treeconv} below. However, for the moment, we ignore
these complications and consider what can be said about the true species tree.

Speciation, which causes branching in a species tree, often occurs by an
\defn{allopatric} process, where a sub-population of organisms is isolated from
the original population by a geographic barrier~\autocite{coyne2004speciation}.
Over time, differing selection pressures on either side of the barrier, coupled
with genetic drift, cause the two populations to diverge genetically. This
eventually results in two distinct species, which will appear as two subtrees
or \defn{clades} in their species tree if descendants of both survive to the
present. Of course, this is not necessarily the case, as selection may cause
one or both of the new sub-populations or their descendants to become extinct.
On the other hand, selection may favour one of the sub-populations, allowing it
to expand into new regions or niches and possibly paving the way for subsequent
speciation events. It must be noted that speciation is not necessarily
allopatric. \defn{Sympatric} speciation refers to the emergence of a new
species without any geographic barrier~\autocite{coyne2004speciation}. Indeed,
there is a whole continuum of possible speciation
processes~\autocite{fitzpatrick2008if}, ranging from completely allopatric to
completely \defn{sympatric}. However, at least for eukaryotic organisms,
allopatric speciation is thought to be by far the most common
type~\autocite{coyne2004speciation}.

A different type of tree used in the study of pathogens is a \defn{transmission
tree}. In such trees, tips represent infected hosts, while internal nodes
correspond to transmissions from one host to another. Transmission trees
generally have branch lengths in units of calendar time, with a root
corresponding to the index case, and branching times indicating times of
transmission. The internal nodes may be labeled with the donor of the
transmission pair, if this is known. The tips of the tree, rather than being
fixed at the present day, are placed at the time at which the individual was
discovered or their pathogen population sampled. Consequently, the transmission
tree may not be ultrametric, but may have tips located at varying distances
from the root. Such trees are said to have \defn{heterochronous} taxa or
samples~\autocite{drummond2003measurably}, in contrast to the
\defn{isochronous} samples found in macro-organisms species trees. A
transmission tree is illustrated in Figure~\ref{fig:contactnet} (right).

When considering epidemics of RNA viruses, transmission trees and species trees
are highly related. The viral population within a single host has often been
referred to as a \defn{quasispecies}~\autocite{domingo2012viral}, and due to
the fast evolutionary rate of these viruses, two hosts' viral populations are
invariably genetically distinct. The transmission process is functionally
identical to allopatric speciation, involving geographic isolation of a
sub-population in a new host and subsequent genetic divergence. However, for
the most part, it is not selection but rather host epidemiology which causes
certain lineages to proliferate while driving others to extinction. A host who
recovers, dies, or is isolated equates to an extinction in the transmission
tree. On the other hand, dense populations in frequent contact often experience
rapid epidemic growth, which causes similarly rapid accumulation of branching
points in the tree.

@@

\subsubsection{Gene trees and viral phylogenies}
\label{subsubsec:genetree}

@@

\subsubsection{Estimating transmission trees}
\label{subsubsec:treeconv}

Most applications of viral phylodynamics, including this project, aim to answer
epidemiological questions~\autocite{volz2013viral}. For example, phylodynamic
methods have been used to investigate the degree of
clustering~\autocite{hughes2009molecular} and the effect of elevated
transmission risk in acute infection~\autocite{volz2012simple} (a more thorough
review is given below in subsection~\ref{subsubsec:appphylo}). For these
applications, we are ultimately interested in the transmission tree, which is a
representation of an epidemiological process (transmission), as opposed to the
viral phylogeny, which represents a biological process (viral evolution). As
discussed above in subsection~\ref{subsubsec:speciestree}, transmission trees and
are in practice unknown, and must be estimated from available data. Here, we
review some of the ways to perform such estimation.

@@

\subsubsection{Tree shapes}
\label{subsubsec:treeshape}

@@

Many tree summary statistics have been developed to assign numerical values to
phylogenies based on their properties. One of the most widely used is Sackin's
index, which measures the imbalance or asymmetry in the tree.

\subsubsection{Applications of phylodynamics}
\label{subsubsec:appphylo}

\subsection{Contact networks}
\label{subsec:contactnet}

\subsubsection{Overview}

@@

\subsubsection{Models for contact networks}
\label{subsubsec:generative}

@@

%\subsection{Approximate Bayesian computation}

@@

\subsection{Approximate Bayesian computation}

\subsubsection{Overview and motivation}

Mathematical models can be grouped into three categories in terms of our
ability to fit them to observed data. First, there is a limited set of models
which can be fit explicitly by using calculus to minimize a loss function. The
canonical example of this type of model is a linear regression, with the
ordinary least-squares loss function. A linear regression is a model of the
relationship between some data points $x_1, \ldots, x_n$ and outcomes $y_1,
\ldots, y_n$ which assumes that the outcomes are linearly related to the data,
modulo some noise caused by measurement errors, environmental fluctuations, and
so on. In other words, there is a constant $b$ such that 
\begin{align}
  \label{eq:regression}
  y_i = bx_i + \varepsilon_i, 
\end{align}
where $\varepsilon_i$ is the error associated with the outcome $y_i$. Notice
that, if we fix $b$, then since $y_i$ and $x_i$ are both given, $\varepsilon_i$
is also fixed. The process of fitting a linear model means choosing the
``best'' $b$. For least squares, this optimal $b$ is defined as the one which
minimizes the sum of the squared $\varepsilon_i$.

@@

\subsubsection{Algorithms for ABC}
\label{subsubsec:abcalg}

@@

% MCMC

>>>>

\section{Phylogenetics and phylodynamics}

\subsection{Phylogenetic trees}

@@

The branches of the tree may have associated lengths, representing either
evolutionary distance or calendar time between ancestors and their descendants.
The term ``evolutionary distance'' is used here imprecisely to mean any sort of
quantitative measure of evolution, such as the number of differences between
the DNA sequences of an ancestor its descendant, or the difference in average
body mass or height. Particular examples of evolutionary distance, based on
genetic data, will be discussed below in subsection~\ref{subsubsec:genetree}. A
phylogeny with branch lengths in calendar time units is often referred to as
\defn{time-scaled}. In a time-scaled phylogeny, the internal nodes can be
mapped onto a timeline by using the tips of the tree, which usually map to the
present day, as a reference point~\autocite{nee1992tempo}. The corresponding
points on the timeline are called \defn{branching times}, and the rate of their
accumulation is referred to as the \defn{branching rate}. Rooted trees whose
tips are all the same distance from the root are called \defn{ultrametric}
trees~\autocite{buneman1974note}. These concepts are illustrated in
Figure~\ref{fig:speciestree}.

@@

\subsection{Species trees and transmission trees}
\label{subsubsec:speciestree}

A \defn{species tree} is a particular type of phylogeny in which the taxa are
species, and the branching times correspond to historical speciation events. By
speciation events, we mean times when two formerly interbreeding groups of
organisms ceased to reproduce with each other. In general, we do not have
access to the true tree relating a particular group of extant species, as this
would usually imply certain knowledge of speciation events in the distant past.
Therefore, all species trees considered by researchers are estimates, based on
the genetic or phenotypic similarity of the extant taxa, the fossil record, or
both. We discuss some of the challenges associated with this estimation in
subsection \ref{subsubsec:treeconv} below. However, for the moment, we ignore
these complications and consider what can be said about the true species tree.

Speciation often occurs by an \defn{allopatric} process, where a sub-population
of organisms is isolated from the original population by a geographic
barrier~\autocite{coyne2004speciation}. Over time, differing selection
pressures on either side of the barrier, coupled with genetic drift, cause the
two populations to diverge genetically. This eventually results in two distinct
species, which will appear as two subtrees or \defn{clades} in their species
tree if descendants of both survive to the present. Speciation is not
necessarily allopatric: \defn{sympatric} speciation refers to the emergence of
a new species without any geographic barrier~\autocite{coyne2004speciation}.
Indeed, there is a whole continuum of possible speciation
processes~\autocite{fitzpatrick2008if}, ranging from completely allopatric to
completely sympatric. However, at least for eukaryotic organisms, allopatric
speciation is thought to be by far the most common
type~\autocite{coyne2004speciation}.

%Of course, this is not
%necessarily the case, as selection may cause one or both of the new
%sub-populations or their descendants to become extinct. On the other hand,
%selection may favour one of the sub-populations, allowing it to expand into new
%regions or niches and possibly paving the way for subsequent speciation events.

A different type of tree used in the study of pathogens is a \defn{transmission
tree}. In such trees, tips represent infected hosts, while internal nodes
correspond to transmissions from one host to another. Transmission trees
generally have branch lengths in units of calendar time, with a root
corresponding to the index case, and branching times indicating times of
transmission. The internal nodes may be labeled with the donor of the
transmission pair, if this is known. The tips of the tree, rather than being
fixed at the present day, are placed at the time at which the pathogen
population of the individual was sampled. Consequently, the transmission tree
may not be ultrametric, but may have tips located at varying distances from the
root. Such trees are said to have \defn{heterochronous} taxa or
samples~\autocite{drummond2003measurably}, in contrast to the
\defn{isochronous} samples found in macro-organisms species trees. A
transmission tree is illustrated in Figure~\ref{fig:contactnet} (A).

\textit{A priori}, transmission trees and species trees have nothing to do with
each other. One represents the evolutionary history of a group of organisms,
the other traces the spread of a disease through a host population. However,
when considering epidemics of RNA viruses, transmission trees and species trees
are highly related. The viral population within a single host has often been
referred to as a \defn{quasispecies}~\autocite{domingo2012viral}, and due to
the fast evolutionary rate of these viruses, two hosts' viral populations are
invariably genetically distinct. The transmission process is functionally
identical to allopatric speciation, involving geographic isolation of a
sub-population in a new host and subsequent genetic divergence. However, for
the most part, it is not selection but rather host epidemiology which causes
certain lineages to proliferate while driving others to extinction. A host who
recovers, dies, or is isolated equates to an extinction in the transmission
tree. On the other hand, dense populations in frequent contact often experience
rapid epidemic growth, which causes similarly rapid accumulation of branching
points in the tree.

@@

\subsection{Gene trees and viral phylogenies}
\label{subsubsec:genetree}

@@

\subsection{Estimating transmission trees}
\label{subsubsec:treeconv}

Most applications of viral phylodynamics, including this project, aim to answer
epidemiological questions~\autocite{pybus2009evolutionary, volz2013viral}. For
example, phylodynamic methods have been used to investigate the degree of
clustering~\autocite{hughes2009molecular} and the effect of elevated
transmission risk in acute infection~\autocite{volz2012simple} (a more thorough
review is given below in subsection~\ref{subsubsec:appphylo}). For these
applications, we are ultimately interested in the transmission tree, which is a
representation of an epidemiological process (transmission), as opposed to the
viral phylogeny, which represents a biological process (viral evolution). As
discussed above in subsection~\ref{subsubsec:speciestree}, transmission trees
and are in practice unknown, and must be estimated from available data. Here,
we review some of the ways to perform such estimation.

@@

\subsection{Tree shapes}
\label{subsubsec:treeshape}

@@

Many tree summary statistics have been developed to assign numerical values to
phylogenies based on their properties. One of the most widely used is Sackin's
index~\cite{shao1990tree}, which measures the imbalance or asymmetry in the
tree.

\subsection{Applications of phylodynamics}
\label{subsubsec:appphylo}

\section{Contact networks}
\label{subsec:contactnet}

\subsection{Overview}

@@

\subsection{Models for contact networks}
\label{subsubsec:generative}

@@

%\section{Approximate Bayesian computation}

@@

\section{Approximate Bayesian computation}

\subsection{Overview and motivation}

\gls{ABC} is the name for a family of techniques which can be used to fit
complex models to observed data. We shall make this precise below, but to fully
describe and motivate \gls{ABC}, it is necessary to first make explicit exactly
what we mean by ``fitting'' a model.

A linear regression is a model of the relationship between some data points
$\vec{x} = (x_1, \ldots, x_n)$ and outcomes $\vec{y} = (y_1, \ldots, y_n)$.
This model assumes that the outcomes are linearly related to the data, modulo
some noise caused by measurement errors, environmental fluctuations, and so on.
In other words, there is a constant $b$ such that $y_i = bx_i + \varepsilon_i$,
where $\varepsilon_i$ is the error associated with the outcome $y_i$. If we
make the additional assumption that the errors are normally distributed, then
the model takes the form
\begin{align}
  y_i = bx_i + \N(0, \sigma^2).
  \label{eq:regression}
\end{align}
When we say that we want to fit this model to observed data, we mean that we
want to find particular values for $b$ and $\sigma$ which are in some way
optimal given $\vec{x}$ and $\vec{y}$. We shall make this precise below. Notice
that, if we fix $b$ and $\sigma$, then we can write down the probability density
of observing $y_i$ given $x_i$. If $f_\N(\cdot \mid \mu, \sigma)$ is the
\gls{pdf} of the normal distribution with mean $\mu$ and variance $\sigma^2$,
and the $y_i$ are all independent, then
\[
  f(\vec{y} \mid \vec{x}, b, \sigma) = 
  \prod_{i=1}^n f_\N(y_i - bx_i \mid 0, \sigma^2).
\]
This gives us a natural criterion for choosing the optimal parameters: we want
to pick $b$ and $\sigma$ which define a model where the probability density of
our observed data is as high as possible. When performing such an optimization,
the above density function is also called the \defn{likelihood} of the
parameters,
\[
  \L(b, \sigma \mid \vec{x}, \vec{y}) = f(\vec{y} \mid \vec{x}, b, \sigma).
\]
The particular $b$ and $\sigma$ which optimize $\L(b, \sigma)$ are called the
\gls{ML} estimates.

%Mathematical models can be grouped into three categories in terms of our
%ability to fit them to observed data. First, there is a limited set of models
%which can be fit explicitly by using calculus to minimize a loss function. The
%canonical example of this type of model is a linear regression. 
%Notice that, if we fix $b$, then since $y_i$ and $x_i$ are both given,
%$\varepsilon_i$ is also fixed. The process of fitting a linear model means
%choosing the ``best'' $b$. For least squares, this optimal $b$ is defined as
%the one which minimizes the sum of the squared $\varepsilon_i$.

@@

\subsection{Algorithms for ABC}
\label{subsubsec:abcalg}

@@

% MCMC
The second method is \gls{ABC}-\gls{MCMC}. This is similar to ordinary Bayesian
\gls{MCMC}, except that a ratio of distances to the observed data replaces the
likelihood ratio. The algorithm begins by sampling a single vector of parameter
values $\theta$ from the prior distribution $\pi(\theta)$. It then proceeds
iteratively: a new parameter vector $\theta^*$ is chosen according to a
proposal distribution $q(\theta^* \mid \theta)$. The proposal distribution $q$
is often taken to be a Gaussian centred at $\theta$. Then $\theta^*$ is
accepted as the new $\theta$ with probability
\[
  \max\left(1, \text{TODO} \right),
\]
or discarded otherwise. This process is iterated until some stopping criterion
is reached, typically a simple limit on the number of steps. After some initial
number of iterations, known as \defn{burn-in}, parameters $\theta$ are
routinely sampled. Since points in parameter space are visited in proportion to
their posterior probability, these samples can be taken to approximate the
posterior distribution on $\theta$, and can be used to calculate point
estimates and confidence intervals.

### 2016-01-29 doc/thesis/objective.tex
More writing

The present study is closely related to three groups of work within
computational epidemiology, which will be discussed in detail below. The first
and largest of these are phylodynamic investigations of epidemiological
parameters, such as transmission rate, recovery rate, and basic reproductive
number. Like our work, these studies make inferences from evolutionary
information in the form of a viral phylogeny, but their aims do not include
uncovering population structure and they typically assume a homogeneously mixed
population. A second, more recent group of studies has evaluated the effect of
network structure on transmission tree shape, by comparing trees simulated from
different network models. These have demonstrated that network structure does
profoundly influence tree shape, but did not attempt to quantitatively infer
the parameter from observed trees. The third group consists of studies which
infer a structural network parameter from namely observed infection and
recovery times using Bayesian methods. These are most similar in aims to our
work, but they rely on epidemiological observations rather than evolutionary
information from phylogenies.

In recent years, a large number of studies have been carried out using
phylodynamic methods to infer epidemiological quantities of interest. The
majority of these employ a Bayesian \gls{MCMC} approach to infer parameters of
an epidemiological model whose likelihood can be calculated, most often some
variation of the birth-death~\autocite{kendall1948generalized} or
coalescent~\autocite{kingman1982coalescent} models. Stadler
\etal~\autocite{stadler2011estimating} develop a formula for the likelihood of
a phylogeny with heterochronous tips under the birth-death model, which has
been used to estimate the basic reproductive number of several viral
epidemics~\autocite{stadler2011estimating}. However, the birth-death model is
cannot tell us anything about population structure, as it assumes that every
individual becomes infected at the same rate. Volz~\autocite{volz2012complex}
writes down the likelihood of a heterochronous phylogeny under a coalescent
model with arbitrarily complex population dynamics. This opens the door to more
complex inferences about population structure, as the population can be
partitioned into compartments with different transmission and recovery rates,
but still assumes that each compartment is homogeneously mixed. In other words, 
the coalescent model can tell us about the \emph{global} structure of a
population, such as whether there exists a high-risk subgroup, but not about
the \emph{local} structure, such as the average number of contacts each
individual has.

The use of models which assume a \defn{panmictic} (that is, homogeneously
mixed) population in phylodynamics has become widespread, so it is natural that
some researchers have investigated the effects of this assumption. In
particular, since phylodynamic methods use phylogenies as their data source,
several studies have examined the shapes of phylogenies arising from
non-panmictic populations. \citeauthor{o2010contact}~(\citeyear{o2010contact})
simulated epidemics over networks with four types of degree distribution. They
then estimated the Bayesian skyride~\autocite{minin2008smooth} population size
trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
incidence and prevalence trajectories, using the method developed by Volz
\etal~\autocite{volz2009phylodynamics}. They found that the concordance between
the two skyrides, as well as the relationship between the skyride and
prevalence curve, was qualitatively different for each degree distribution.
\citeauthor{leventhal2012inferring} simulated transmission trees over \gls{ER},
\gls{WS}, \gls{BA}, and full networks with fixed number of nodes and mean degree.
They calculated Sackin's index of the simulated trees while varying the
epidemic, network, and sampling parameters, and found that the relationship
between these parameters and Sackin's index varies considerably among the
different network models.

The third group of studies was pioneered by Britton and
O'Niell~\autocite{britton2002bayesian}, who developed a Bayesian method to
infer the $p$ parameter of an \gls{ER} network, along with the $\beta$ and
$\gamma$ parameters of \gls{SI} model. Their method is able to use either
infection and removal times, or removal times only. However, it is designed for
only a small number of observations (15 and 42 cases in their applications),
and their estimates of $p$ for real outbreaks were mostly uninformative (95\%
confidence intervals [0.11-0.96] and 0.055-0.96]; $p$ is bounded in [0, 1]).

>>>>

The present study is closely related to three groups of work with distinct
objectives, which will be reviewed in detail below. 
%, by
%comparing trees simulated from different network models. These have
%demonstrated that network structure does profoundly influence tree shape, but
%did not attempt to quantitatively infer the parameter from observed trees. The
%third group consists of studies which infer a structural network parameter from
%observed infection and recovery times, using Bayesian methods. These are most
%similar in aims to our work, but they rely on epidemiological observations
%rather than evolutionary information from phylogenies.

The first and largest group of related studies are phylodynamic investigations
of epidemiological parameters such as transmission rate, recovery rate, and
basic reproductive number~\autocite{pybus2009evolutionary, volz2013viral}. Like
our work, these studies make inferences about epidemiological processes from
the genetic diversity of virus populations, which is usually represented in the
form of a phylogeny. The majority of these employ a Bayesian \gls{MCMC}
approach to infer parameters of an epidemiological model whose likelihood can
be calculated, most often some variation of the
birth-death~\autocite{kendall1948generalized} or
coalescent~\autocite{kingman1982coalescent} models. Stadler
\etal~\autocite{stadler2011estimating} develop a formula for the likelihood of
a phylogeny with heterochronous tips under the birth-death model, which has
been used to estimate the basic reproductive number of several viral
epidemics~\autocite{stadler2011estimating}. However, the birth-death model is
cannot tell us anything about population structure, as it assumes that every
individual becomes infected at the same rate. Volz~\autocite{volz2012complex}
writes down the likelihood of a heterochronous phylogeny under a coalescent
model with arbitrarily complex population dynamics. This opens the door to more
complex inferences about population structure, as the population can be
partitioned into compartments with different transmission and recovery rates,
but still assumes that each compartment is homogeneously mixed. In other words,
the coalescent model can tell us about the \emph{global} structure of a
population, such as whether there exists a high-risk subgroup, but not about
the \emph{local} structure, such as the average number of contacts each
individual has.

A second, more recent group of studies has evaluated the effect of network
structure on transmission tree shape. The use of models which assume a
\defn{panmictic} (that is, homogeneously mixed) population in phylodynamics has
become widespread, so it is natural that some researchers have investigated the
effects of this assumption. In particular, since phylodynamic methods use
phylogenies as their data source, several studies have examined the shapes of
phylogenies arising from non-panmictic populations. 
\citeauthor{o2010contact}~(\citeyear{o2010contact}) simulated epidemics over
networks with four types of degree distribution. They then estimated the
Bayesian skyride~\autocite{minin2008smooth} population size trajectory in two
ways: from the phylogeny, using \gls{MCMC}; and from the incidence and
prevalence trajectories, using the method developed by Volz
\etal~\autocite{volz2009phylodynamics}. They found that the concordance between
the two skyrides, as well as the relationship between the skyride and
prevalence curve, was qualitatively different for each degree distribution.
\citeauthor{leventhal2012inferring} simulated transmission trees over \gls{ER},
\gls{WS}, \gls{BA}, and full networks with fixed number of nodes and mean
degree. They calculated Sackin's index of the simulated trees while varying the
epidemic, network, and sampling parameters, and found that the relationship
between these parameters and Sackin's index varies considerably among the
different network models. In summary, studies in this group have demonstrated
that network structure profoundly influences tree shape, but have not attempted
to quantitatively infer network parameters from observed trees.

Finally, a third group of studies has used Bayesian methods to infer a
structural network parameter from observed infection and recovery times. This
third group is most similar in aims to our own work; rather than using
epidemiological observations, we employ viral phylogenies. The pioneering study
in this group was by
\citeauthor{britton2002bayesian}~(\citeyear{britton2002bayesian}), who
developed a Bayesian method to infer the $p$ parameter of an \gls{ER} network,
along with the $\beta$ and $\gamma$ parameters of \gls{SI} model. Their method
is able to use either infection and removal times, or removal times only.
However, it is designed for only a small number of observations (15 and 42
cases in their applications), and their estimates of $p$ for real outbreaks
were mostly uninformative (95\% confidence intervals [0.11-0.96] and
0.055-0.96]; $p$ is bounded in [0, 1]).
\citeauthor{groendyke2011bayesian}~(\citeyear{groendyke2011bayesian})
significantly updated and extended the methodology of
\citeauthor{britton2002bayesian}, and applied it to a measles outbreak
affecting 188 individuals. They were able to obtain a much more informative
estimate of $p$, although this data set included both symptom onset and
recovery times for all individuals. They also assumed that the entire network
was infected, so that the network size was exactly 188. It is unclear how well
their method would scale to networks with thousands of nodes.

### 2016-01-29 doc/thesis/results.tex
More writing

\subsection{Estimable parameters}

@@

The projections show that all three values are well separated from each other
in many of the scenarios. With smaller trees, it becomes harder to disinguish
$\alpha = 0.5$ from $\alpha = 1.0$ in two dimensions.

\begin{figure}[ht]
  \centering
  \label{fig:pakpca}
  \includegraphics{kernel_pa_kpca}
  \caption[Projection of kernel matrix for different attachment power values
  onto its first two principal components]{
    Projection of the kernel matrix for different preferential attachment power
    values onto its first two principal components, for eight simulation
    scenarios. Each point corresponds to a simulated transmission tree, and is
    coloured by preferential attachment power. Facets are number of infected
    nodes (horizontal), and number of sampled tips (vertical). The parameters
    to the tree kernel were $\lambda = 0.3$ and $\sigma = 5$, and the nLTT was
    not used. Qualitatively, trees with a larger number of tips are easier to
    separate in kernel space, regardless of what sampling proportion they
    represent. In all cases, the highest attachment power can be separated from
    the other two, but the two lowest values become hand to distinguish with in
    the 100-tip trees.
  }
\end{figure}

\subsection{Accuracy and precision}

\begin{figure}[ht]
  \centering
  %\includegraphics{gridsearch_pa_kernel}
  \caption{Grid search kernel scores.}
  \label{fig:gridkernel}
\end{figure}

>>>>

\subsection{Separable parameters}

Our first simulation-based experiment was designed to determine which network
model parameters were potentially estimable by kernel-ABC. To do this, we
simulated trees under three distinct values of each parameter of interest, and
evaluated the accuracy of a kernel-\gls{SVM} classifier for the parameter. This
is the same approach taken by \textcite{poon2015phylodynamic}, and has the dual
purpose of allowing us to identify the optimal kernel meta-parameters
\gls{sigma} and \gls{lambda}. The parameters of the \gls{ER}, \gls{WS}, and
\gls{BA} network models were examined, but we will discuss only the results for
the $\alpha$ and prevalence parameters of the \gls{BA} model in the main text.
The results for the remaining models and their parameters can be found in the
supplemental material.

In Figure~\ref{fig:patrees}, we show that trees simulated under different
values of \gls{alpha} are visibly quite distinctive. In particular, higher
values of \gls{alpha} result in networks with a small number of highly
connected nodes (see Figure~\ref{fig:pabounds}) which, once infected, are
likely to transmit to many other nodes. This results in a more unbalanced,
ladder-like structure in the phylogeny. Kernel-\gls{PCA} projections show that
all three \gls{alpha} values are well separated from each other under several
different sampling scenarios (Figure~\ref{fig:pakpca}). With smaller trees,
it becomes harder to visually disinguish \gls{alpha} = 0.5 from \gls{alpha} =
1.0 using the first two principal components.

With suitably chosen meta-parameters, the accuracy of the kernel-\gls{SVM}
classifier for \gls{alpha} was very high under a variety of prevalence and
sampling scenarios (Figure~\ref{fig:pacrossv}). Accuracy was highest, with
$R^2$ values above 0.95, for the largest trees and complete sampling (bottom
center panel). However, even for trees of size 100 sampled from an epidemic on
2000 nodes, the $R^2$ was above 0.8 (top right panel). In all cases, the
accuracy of a Sackin's index-based classifier was also quite high, at about
0.75. The fact that Sackin's index is informative of \gls{alpha} was
unsurprising, given the clear differences in tree balance observed under
different \gls{alpha} values. The \gls{nltt}
statistic~\autocite{janzen2015approximate} did not improve the classifier, but
rather slightly reduced the $R^2$ in most cases.

\begin{figure}[ht]
  \centering
  \label{fig:patrees}
  \includegraphics{kernel_pa_trees}
  \caption[Visibly distinctive trees simulated under three values of \gls{alpha}]{
    Epidemics of 1000 infected were simulated on \gls{BA} networks of 5000
    nodes, with \gls{alpha} equal to 0.5, 1.0, or 1.5. Transmission trees were
    created by sampling 500 infected nodes. Higher \gls{alpha} values produce
    networks with a small number of highly-connected nodes, which results in a
    highly unbalanced, ladder-like tree structure.
  }
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \label{fig:pakpca}
  \includegraphics{kernel_pa_kpca}
  \caption[Projection of kernel matrix for different attachment power values
  onto its first two principal components]{
    Projection of the kernel matrix for different preferential attachment power
    values onto its first two principal components, for eight simulation
    scenarios. Each point corresponds to a simulated transmission tree, and is
    coloured by preferential attachment power. Facets are number of infected
    nodes (horizontal), and number of sampled tips (vertical). The parameters
    to the tree kernel were $\lambda = 0.3$ and $\sigma = 4$, and the nLTT was
    not used. Qualitatively, trees with a larger number of tips are easier to
    separate in kernel space, regardless of what sampling proportion they
    represent. In all cases, the highest attachment power can be separated from
    the other two, but the two lowest values become hand to distinguish with in
    the 100-tip trees.
  }
\end{figure}

We also considered the possibility of inferring the number of infected nodes,
or \defn{prevalence}, under this model. All parameters except \gls{I} were
fixed at the following values: \gls{N} = 5000, \gls{alpha} = 1.0, and \gls{m} =
2. As shown in Figure~\ref{fig:prevtrees}, the prevalence had no obvious effect
on the tree shape. However, a kernel-\gls{SVM} classifier was able to
distinguish the number of infected nodes with high accuracy ($R^2 > 0.9$;
Figure~\ref{fig:prevcrossv}). Moreover, the use of the \gls{nltt} statistic
improved classification accuracy by a small amount, in contrast to the results
for \gls{alpha}. In contrast, a Sackin's index-based classifier displayed
extremely poor performance ($R^2$ < 0.1, not shown).

It is important to note that, if we cut off the epidemic simulation when 500
nodes are infected, the resulting tree will be shorter (in calendar time) than
if we continue until 2000 nodes are infected. However, this information is not
used when building the classifier, since the branch lengths in each tree are
scaled by their mean. Therefore, the high performance of the classifier is due
to structural differences captured by the tree kernel, rather than the trees
simply having different heights.

\begin{figure}[ht]
  \centering
  \label{fig:prevtrees}
  \includegraphics{kernel_nsimnode_trees}
  \caption[Similarly shaped simulated under three values of \gls{I}]{
    Epidemics were simulated on networks of size 5000 until \gls{I} = 500,
    1000, or 2000 nodes were infected. When scaled to the same height, there is
    no immediately visible distinction in shape between trees simulated under
    different \gls{I} values.
  }
\end{figure}

\begin{figure}[ht]
  \centering
  \label{fig:prevcrossv}
  \includegraphics{kernel_nsimnode_crossv}
  \caption[Cross-validation performance of kernel-\gls{SVM} classifier for
           prevalence]{
    For \gls{BA} networks of size \gls{N} = 5000, a kernel-\gls{SVM} classifier
    is very accurate at predicting \gls{I}. Epidemics were simulated until
    \gls{I} = 500, 1000, or 2000 nodes were infected, and either 100 or 500
    nodes were sampled for inclusion in a transmission tree. The parameters of
    the \gls{BA} network were \gls{alpha} = 1.0 and \gls{m} = 2. A Sackin's
    index-based classifier performed very poorly on these data ($R^2 = $ TODO
    for 100-tip trees and blah for 500-tip trees, not shown).
  }
\end{figure}

\subsection{Accuracy of marginal estimates}

We used grid search to obtain \defn{marginal} estimates for each network
parameter while holding all other parameters fixed. We observed that kernel
scores were highest at the values of \gls{alpha} on the grid closest to the
true values, as shown in Figure~\ref{fig:gridkernel}. However, there was a much
stronger spike in kernel scores near the true value for \gls{alpha} = 1.0 and
1.25. This is recapitulated when we look at the accuracy of point estimates
obtained by taking the grid value with the highest median kernel score. As
shown in Figure~\ref{fig:gridest}, while the estimates are generally close to
the true value, they are much closer for \gls{alpha} = 1.25 than for the other
values.

\begin{figure}[ht]
  \centering
  \includegraphics{gridsearch_pa_kernel}
  \caption[Grid search kernel scores]{
    Grid search kernel scores for testing trees simulated under various
    \gls{alpha} values. All epidemics had \gls{I} = 1000 infected nodes, on
    \gls{BA} networks of size \gls{N} = 5000 with \gls{m} fixed at 2. Colours
    indicate the number of sampled tips.
  }
  \label{fig:gridkernel}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics{gridsearch_pa_estimates}
  \caption[Marginal estimates of \gls{alpha} obtained with grid search]{
    Marginal estimates of \gls{alpha} obtained with grid search. Training trees
    were simulated on a narrowly spaced grid of \gls{alpha} values, and
    compared to testing trees using the tree kernel. The \gls{alpha} value in
    the grid with the highest median kernel score was taken as the point
    estimate for the testing tree. These point estimates are shown as black
    dots. The dashed line is the identity.
  }
  \label{fig:gridest}
\end{figure}

\subsection{Accuracy of estimates with full ABC}
